%%
%% TOMS Algorithm Article
%% Manuscript format for review
%%
\documentclass[manuscript,screen,review]{acmart}

%% Journal information for TOMS
\setcopyright{rightsretained}
\acmJournal{TOMS}
\acmYear{2026}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmMonth{1}
\acmDOI{10.1145/XXXXXXX}

%% Packages
\usepackage{listings}
\usepackage[T1]{fontenc}


\lstdefinestyle{tightcode}{
  language=C,
  basicstyle=\ttfamily\small,
  columns=fixed,
  basewidth=0.55em,   % horizontal spacing between characters
  lineskip=-2pt,      % vertical spacing between lines
  aboveskip=1em,
  frame=single,
  framerule=0.5pt,
  framesep=3pt,
  xleftmargin=4em,
  xrightmargin=4em
}

%% BibTeX command
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Bibliography
\bibliographystyle{ACM-Reference-Format}

\begin{document}

%%
%% Title
%%
\title{Algorithm XXX: Randompack, A Portable C Library for Reproducible Random Number Generation}

%%
%% Author information - FILL IN YOUR DETAILS
%%
\author{Kristján Jónasson}
\authornote{Corresponding author.}
\email{jonasson@hi.is}
\orcid{0000-0000-0000-0000}
\affiliation{%
  \institution{University of Iceland}
  \city{Reykjavík}
  \country{Iceland}
}


%%
%% Abstract
%%
\begin{abstract}
\end{abstract}

%%
%% Keywords
%%
\keywords{}

\maketitle

\section{Introduction}

The computer generation of pseudo-random numbers has a history dating
back to the first computers, with von Neumann's middle-square idea
(1946) \cite{vonneumann1951rejection,lecuyer2017history} and Lehmer's
linear congruential generator (LCG, 1949)
\cite{lehmer1951mathematical,lecuyer2017history}. The first generators
were aimed at generating random integers (typically in $[0, 2^k)$) and
uniformly distributed reals in $[0,1]$ (through division by $2^k$), but
soon other distributions followed. In fact von Neumann's paper
\cite{vonneumann1951rejection} had a clever algorithm for drawing
exponentially distributed samples, also obtainable by taking the
logarithm of uniform samples. An early and seminal method for random
normals is given by Box and Muller \cite{box1958normal}. Other important
20th century milestones are the Tausworthe random number generators
(RNGs, 1965) \cite{tausworthe1965linear}, the Wichmann-Hill RNG (1982)
\cite{wichmann1982efficient}, the Park-Miller LCG (1988)
\cite{park1988random}, RANLUX (1994) \cite{luscher1994portable}, adopted
in the CERN ROOT framework, and the Mersenne Twister (1998)
\cite{matsumoto1998mersenne} with its extremely long period,
$2^{19937}-1$.

Parallel to the uniform generator development new methods for
generating non-uniform variates were developed. The normal
distribution plays an important role, and other important ones are the
exponential and the gamma distributions. Random sampling from several
other distributions can be done by combining samples from these three
(see Section xxx). The Box-Muller method (1958) requires the
evaluation of a square root, a logarithm, a sine, and a cosine for each
pair of samples, but a few years later Marsaglia and Bray proposed the
polar method where the two trigonometrict functions were replaced by
rejection sampling with acceptance rate $\pi/4$. Rejection sampling
had already been used by von Neuman in his exponential method
\cite{vonneumann1951rejection}. It also plays a central role in the
ratio of uniforms method (1977) \cite{kinderman1977ratio,leva1992fast}
which has an acceptance rate of 73\% and needs two uniform samples but
only 0.012 logarithms and no square roots per normal sample. The next
advance, also using rejection, was the ziggurat method (1984, 2000)
proposed in two stages by Marsaglia and Tsang
\cite{marsaglia1984fast,marsaglia2000ziggurat}, the revised method
drawing from the tail using an earlier idea \cite{marsaglia1964tail}.
The seminal article from 2000 gives ziggurat algorithms for both
normal and exponential variates, as well as a notably elegant method
for the gamma distribution, that has since been adopted by most RNG
libraries. All three are implemented in Randompack. Before 2000
various algorithms had been devised for the gamma distribution, all of
them based on some sort of rejection and many quite complex.

Around the turn of the century, random number research placed
increasing emphasis on the statistical quality of RNGs, on devising
methods for creating independent streams of random numbers, and on
ensuring long periods, spurred by faster, often parallel computers.
For example, on a modern computer, the output of Park–Miller and other
32-bit-state RNGs repeats itself after only a few seconds on a single
core of a modern computer. This development increased the importance
of empirical testing. Knuth had already described rigorous statistical
tests of RNGs in 1969 in \emph{The Art of Computer Programming}
\cite{knuth1969art}, but the next major milestones were the Diehard
suite in 1996 \cite{marsaglia1996diehard}, DieHarder in 2006
\cite{Brown2006DieHarder}, TestU01 in 2007 \cite{lecuyer2007testu01}
with its well-known SmallCrush, Crush, and BigCrush batteries, and
PractRand \cite{dotyhumphrey2010pract}. TestU01 exposed weaknesses
in many widely used RNGs, which failed its crush-batteries. More
recently, Foreman et al. \cite{foreman2024statistical} described an
openly available RNG testing environment built on existing suites. In
parallel with these developments, standards agenecies on both sides of
the Atlantic developed methods for testing cryptographic RNGs
\cite{schindler2002evaluation,rukhin2010nist,saarinen2022sp}.

We proceed with the summary of RNG development where we left off with
the widely used Mersenne Twister (MT; Table~\ref{tab:language-rngs}).
One of the first RNG to offer independent streams, paving the road for
the subsequent advances was MRG32k3a
\cite{lecuyer1999good,lecuyer2002streams}, with a period of $2^{191}$.
The WELL generators \cite{panneton2006well} were intended to (and did)
improve on the MT statistical quality although they did not obtain
widespread adoption in software. The oposite can be said of the RNGs
sometimes called the xor family, pinoeered by Marsaglia in 2003 with
xorshift and XORWOW \cite{marsaglia2003xorshift}, and further developed,
primarily by Vigna, in a series of articles
\citep{panneton2005xorshift,vigna2016experimental,vigna2017further,
  blackman2021scrambled}. The current flagships are the closely related
xoshiro256++ and xoshiro256**, that have been adopted as default
generators by Rust, Julia, C\#, and gfortran
(Table~\ref{tab:language-rngs}). All generators in the xor family
support so-called \emph{jumping}, whereby the state can be advanced
efficiently by very large amounts, such as $2^{128}$ steps, to create
independent streams. The aforementioned Practrand suite included an RNG,
sfc64, that also allows independent streams. Interestingly, sfc64 is
planned to become the default random number generator in a forthcoming
release of the Los Alamos MCNP radiation-transport code.
\cite{josey2023reassessing,gibson2025mcnpnewsletterq1}. 

Some important pseudorandom number generators (PRNGs) emerged from the
search for fast software stream ciphers. Such ciphers can also serve as
cryptographically secure PRNGs (CSPRNGs), notably ChaCha20 (2008)
\cite{bernstein2008chacha}. Another important milestone came in 2011
with the Random123 library by Salmon and Moraes
\cite{salmon2011parallel}. Its counter-based Threefry and Philox
generators have become popular for simulation on GPUs and are the
default in several systems (Table~\ref{tab:language-rngs}). They allow
the generation of a vast number of independent streams.

Another important development was O'Neill's PCG family (2014)
\cite{oneill2014pcg}. It also supports many streams via a
stream-selector termed \emph{increment}. A 64-bit version of it has
been adopted as NumPy's default RNG engine. In 2019 O'Neill suggested
an improved version, PCG64-DXSM, which is planned to become default
for NumPy \cite{oneill2019pcg64dxsm-post,numpy_pcg64dxsm_upgrade}. One
more way of getting many streams is via stream splitting, an
object-oriented approach implemented in SplitMix (2014)
\cite{steele2014splitmix} and refined in the LXM family (2021)
\cite{steele2021lxm}. These RNGs provide both \emph{generate} and
\emph{split} as primitive operations, the latter creating two
independent generators. Java's default RNG is from the LXM family,
while SplitMix has primarily found application in spreading an initial
seed across the state of other generators.

Finally it is worth noting that in 2017 L'Ecuyer wrote a comprehensive
history of uniform pseudo-random number generators
\cite{lecuyer2017history}. The earlier article by James (1990)
\cite{james1990review}, as well as the TestU01 article
\cite{lecuyer2007testu01}, also provide useful historical overviews.
More recent, though in both cases limited, reports on RNG testing are
given by \cite{bhattacharjee2022search} and
\cite{josey2023reassessing}.

\begin{table}
\caption{Default random number generators in selected languages and
  libraries. The generators are discussed briefly in the main text.}
\label{tab:language-rngs}
\centering
\begin{minipage}{0.52\columnwidth}
\centering
\begin{tabular}{l@{\hspace{2.5em}}l}
\toprule
Language/library          & Default engine \\
\midrule
Go                        & ChaCha8 \\
Rust (rand default)       & ChaCha12 \\
Java                      & LXM (L32X64MixRandom) \\
Boost                     & Mersenne Twister\textsuperscript{a} \\
C++ standard library      & Mersenne Twister\textsuperscript{a} \\
Intel oneMKL              & Mersenne Twister\textsuperscript{a} \\
Gnu Scientific Library    & Mersenne Twister \\
Matlab (core)             & Mersenne Twister \\
Python random module      & Mersenne Twister \\
PyTorch (CPU)             & Mersenne Twister \\
Ruby                      & Mersenne Twister \\
R (base)                  & Mersenne Twister \\
R (parallel)              & MRG32k3a \\
R (dqrng)                 & xoroshiro (128++) \\
NumPy                     & PCG64 (XSL-RR 128/64) \\
PyTorch (GPU)             & Philox \\
TensorFlow                & Philox \\
Google JAX                & Threefry \\
Matlab (parallel toolbox) & Threefry \\
CUDA cuRAND               & XORWOW \\
JavaScript (V8)           & xorshift (128+) \\
C\#                       & xoshiro (256**) \\
gfortran                  & xoshiro (256**) \\
Julia                     & xoshiro (256++) \\
Rust (rand SmallRng)      & xoshiro (256++) \\
\bottomrule
\end{tabular}
\par\medskip
\footnotesize
\raggedright
\textsuperscript{a}No unique default RNG exists; the table reports the
engine most commonly used in the official documentation and examples.
\end{minipage}
\end{table}

% Rest of introduction

	% •	motivate the problem
	% •	place the work historically
	% •	state what Randompack provides
	% •	explain what is new or distinctive
	% •	outline the rest of the paper


\section{Design and interface}
\subsection{Design goals}
Randompack is intended to fill a gap in the random-number-generation
support available in programming languages and libraries by providing a
C library with a selection of modern, high quality, and efficient
pseudo-random number generators, together with a range of continuous and
discrete distributions. The C libraries that come closest are GSL,
UNURAN \cite{hormann2000randomnumber}, the NAG library, and Intel
oneMKL. However none of these libraries are permissively licensed and
only the first two are open source. Moreover, those two are somewhat
dated and rely largely on older underlying RNG engines.

C is a particularly suitable implementation language because it can be
interfaced easily from many other programming languages. As an added
benefit, this makes it possible to obtain not only exact repeatability
across computers, as for example in NumPy, but also exact repeatability
across programming languages. The importance of repeatability as a basic
requirement for high-quality random number generation has been stressed
by L’Ecuyer \cite{lecuyer2015rng,lecuyer2017parallel}.

Randompack is intended for scientific and statistical computing rather
than cryptographic use. Its design emphasizes high-throughput
generation of pseudo-random numbers for Monte Carlo simulation,
bootstrap resampling, stochastic optimization, and related large-scale
computational tasks in which both speed and statistical quality
matter. A further design goal is modularity: It should be reasonably
easy to add engines and distributions to allow using Randompack
as a test bed for new ideas.

Further design goals include a small, stable C interface, portable
implementation, thread-safety, few external dependencies, clear
separation between engine and distribution layers, and as indicated
above, support for independent substreams, and bit-level
reproducibility across platforms.

\subsection{User interface}
The primary entry points are the constructor function
\texttt{randompack\_create}, several sampling functions (one for each
supported distribution), and several seeding, setup, and state control
functions. The constructor returns a handle to an opaque structure
representing a random number generator (RNG). This object contains the
underlying engine, its state, and other internal information needed by
the library.

The following minimal example creates an RNG with the default engine,
obtained by passing \texttt{0} to \texttt{randompack\_create}. Then five
samples from the standard normal distribution are drawn and printed out.

\begin{lstlisting}[style=tightcode]
#include <stdio.h>
#include "randompack.h"
int main(void) {
  randompack_rng *rng = randompack_create(0);
  double x[5];
  randompack_norm(x, 5, rng);
  for (int i = 0; i < 5; i++) printf("%g\n", x[i]);
  randompack_free(rng);
  return 0;
}
\end{lstlisting}

A specific engine can be selected by name, e.g. 
\texttt{randompack\_create(\"ranlbux++\")}. To seed the RNG after
creation use for example \texttt{randompack\_seed(42, 0, rng)}. The
second argument is a spawn key (0 means none is specified).

\subsection{Implemented engines}
Randompack implements 11 different PRNG random generators or engines.
All these date from the last 20 years. The primary inclusion criteria
for the engines are that they should

\begin{enumerate}
\item have period at least $2^{64}$,
\item have a compact state size,
\item pass rigorous statistical tests, including BigCrush,
\item be computationally efficient, and
\item support the construction of independent substreams.
\end{enumerate}

\newcommand{\scrambledandweb}{\cite{blackman2021scrambled,vigna2026shootout}}
\newcommand{\oneillpcgandpost}{\cite{oneill2014pcg,oneill2019pcg64dxsm-post}}

\begin{table}
\caption{Random number generators (engines) implemented in Randompack.
The default generator, a SIMD-accelerated version of the xoshiro256++
with short name x256++simd, is not listed.}
\label{tab:engines}
\begin{tabular}{llcclcc}
\toprule
engine & short name & state words & period\textsuperscript{a} & authors & year & references \\
\midrule
xoshiro256++   & x256++   & 4 & $2^{256}-1$        & Vigna and Blackman  & 2019 & \scrambledandweb\\
xoshiro256**   & x256**   & 4 & $2^{256}-1$        & Vigna and Blackman  & 2018 & \scrambledandweb\\
xorshift128+   & x128+    & 2 & $2^{128}-1$        & Vigna               & 2014 & \cite{vigna2016experimental}\\
xoroshiro128++ & xoro++   & 2 & $2^{128}-1$        & Vigna and Blackman  & 2019 & \scrambledandweb\\
PCG64 DXSM     & pcg64    & 4 & $2^{128}$          & O'Neil              & 2019 & \oneillpcgandpost\\
squares64      & squares  & 2 & $2^{64}$           & Widynski            & 2021 & \cite{widynski2020squares}\\
Philox-4x64    & philox   & 6 & $2^{256}$          & Salmon and Moraes   & 2011 & \cite{salmon2011parallel}\\
sfc64          & sfc64    & 4 & $\approx 2^{255}$  & Chris Doty-Humphrey & 2010 & \cite{dotyhumphrey2010pract}\\
cwg128         & cwg128   & 8 & $2^{128}$          & Działa              & 2022 & \cite{dziala2023collatz}\\
ranlux++       & ranlux++ & 9 & $\approx 2^{570}$  & Sibidanov           & 2017 & \cite{sibidanov2017ranluxpp}\\
ChaCha20       & chacha20 & 6 & $2^{35}$           & Bernstein           & 2008 & \cite{bernstein2008chacha}\\
\bottomrule
\end{tabular}
\par\medskip
\centering
\begin{minipage}{\dimexpr\linewidth-2cm\relax}
\footnotesize
\raggedright
\textsuperscript{a}For PCG64 DXSM and cwg128 the listed period is for
each increment (out of $2^{127}$), for Philox and squares64 it is for
each key (out of $2^{128}$ and $2^{64}$, respectively, and for
ChaCha20 it is for each nonce, (out of $2^{96}$).
\end{minipage}
\end{table}

Table \ref{tab:engines} lists the selected engines. Most of them have
already been mentioned in the Introduction, the exceptions being the
recent squares64, a fast counter-based RNG and, cwg128, also recent,
simple and stream-friendly. A few of the generators in the table do
not fulfill all the inclusion criteria: xorshift128+ and
xoroshiro128++ do not reliably pass BigCrush, but they are very
efficient, do appear in Table \ref{tab:language-rngs}, and are
important in the evolution of the xor family. ChaCha20 is not
particularly efficient and its period is relatively short, but it is
included to provide one cryptographically secure engine.

While many of the engines from Table \ref{tab:language-rngs} are
implemented in Randompack, there are some that are absent. Among those
is the most popular one, the Mersenne Twister. Its state is 312 64-bit
words, 35 times more than the largest Randompack engine (ranlux++).
Moreover it fails BigCrush. Of the remaining generators, XORWOW does not
pass BigCrush reliably, but three others, Threefry, Java's LXM
generator, and MRG32k3a, are stronger candidates for later inclusion.
They are high-quality but probably somewhat slower than most of those
present.

With one exception, the C code for the engines is copied directly from
their original or standard public description, with minor adaptation to
the Randompack setting. The exception is ranlux++, a modernized version
of ranlux due to Sibidanov (2017) which was rewritten from his
mathematical description because all available implementations,
including Sibidanov's, are Gnu-licensed
\cite{sibidanov2017ranluxpp,hahnfeld2021ranluxpp,jirka2021portable}.

\subsection{State initialization and stream control}
Randompack provides four classes of of RNG engine initialization:
deterministic seeding, explicit stream selection, randomization from
system entropy, and exact assignment of the full internal state.

Seeding deserves careful treatment as some RNGs may show poor
statistical behaviour if seeded directly with a small seed. A better
approach is to mix the seed and spread it across the full state, using a
process that is radically different from the RNG being seeded
\cite{matsumoto2007initialization,oneill2014pcg,blackman2021scrambled}.
As noted in the Introduction, SplitMix \cite{steele2014splitmix} could
be used for this purpose, but O'Neill's \texttt{seed\_seq\_fe128}
\cite{oneill2015seedseq} provides an alternative adopted in Randompack.
It hashes and mixes a vector of seeds into a 128-bit \emph{entropy pool}
and then applying different hash functions to the pool to generate the
output state. O'Neill aims for strict avalanche, meaning that flipping a
bit in the seed material flips each output bit with probability $1/2$,
and reports empirical results consistent with such behaviour. NumPy's
seeding is likewise based on O'Neill's \texttt{seed_seq} work.

In Randompack the user seed together with a spawn key of arbitrary
length constitutes the input to the seeding mixer. This creates one
way of creating an arbitrary number of deterministically seeded
substreams. The spawn key could, for instance, consist of MPI rank,
thread id, task number, etc. For Randompack's engines the probability
that two such substreams overlap will be vanishingly small, but, if
the user prefers, Randompack also supports the built-in features of
the different engines for creating mathematically proven independent
streams: jumps for ranlux and the xor family, keys for the
counter-based engines, $(a,b,c)$ state for sfc64, increments for pcg64
and cwg128, and the nonce for ChaCha20. These stream-selection
operations would typically be used together with the duplicate
function when setting up parallel simulations.

Randompack’s randomize function obtains cryptographically secure system
entropy from the operating system: it uses \texttt{RtlGenRandom} on
Windows, \texttt{arc4random\_buf} on macOS and OpenBSD, and
\texttt{getrandom} on Linux and other Unixes which support it, with
\texttt{/dev/urandom} as a fallback. If that too is unavailable, an
emergency fallback mixes in the system clock. 

\cite{matsumoto2007initialization}

%    2.3 Engines, streams, and seeding
% 3.	Algorithms and implementation
%    3.1 Uniforms and bounded integers
%    3.2 Exponential, normal, gamma, and related distributions
%    3.3 Permutations and sampling without replacement
%    3.4 SIMD and performance-oriented implementation
% 4.	Discussion
% 5.	Conclusioge
\section{Design and interface}

\section{Algorithms and implementation}

\section{Discussion}

%%
%% Bibliography
%%
\bibliography{randompack}

%%
%% Appendix (optional)
%%
\appendix

\section{Algorithm Listings}

\section{Additional Performance Data}

\end{document}
